---
title: "ENV 790.30 - Time Series Analysis for Energy Data | Spring 2021"
subtitle: "Assignment 6 - Due date 03/16/22"
author: "Soman Ul Haq"
output: pdf_document
geometry: margin=2.54cm
header-includes:
  - \usepackage{enumerate}
  - \usepackage{enumitem}
editor_options: 
  chunk_output_type: console
---
  
## Directions
  
  You should open the .rmd file corresponding to this assignment on RStudio. The file is available on our class repository on Github. And to do so you will need to fork our repository and link it to your RStudio.  

Once you have the project open the first thing you will do is change "Student Name" on line 3 with your name. 
Then you will start working through the assignment by  **creating code and output** that answer each question. Be sure to use this assignment document. Your report should contain the answer to each question and any plots/tables you obtained (when applicable).

When you have completed the assignment, **Knit** the text and code into a single PDF file. Rename the pdf file such that it includes your first and last name (e.g., "LuanaLima_TSA_A06_Sp22.Rmd"). Submit this pdf using Sakai. 

## Questions 

This assignment has general questions about ARIMA Models. 

Packages needed for this assignment: "forecast","tseries". Do not forget to load them before running your script, since they are NOT default packages.\\

```{r packages}
#Load/install required package here
library(lubridate)
library(ggplot2)
library(forecast)  
library(Kendall)
library(tseries)
library(knitr)
library(sarima)


```


## Q1
Describe the important characteristics of the sample autocorrelation function (ACF) plot and the partial sample autocorrelation function (PACF) plot for the following models: 

\begin{enumerate}[label=(\alph*)]

\item AR(2)

> Answer: An AR process of order 1 will have these types of ACF and PACF: the ACF will decay exponentially with time whereas the PACF will be used to identify the order depending on where it cuts off. In this case, the PACF will cut off after the first two high bars.

\item MA(1)

> Answer: A Moving Average process of order 2 will have the following types of ACF and PACF: the ACF will show the order of the process depending on where it cuts off, and the PACF will decay exponentially and slowly. In this case, the ACF will cut off after the first high bar. 

\end{enumerate}

## Q2
Recall that the non-seasonal ARIMA is described by three parameters ARIMA$(p,d,q)$ where $p$ is the order of the autoregressive component, $d$ is the number of times the series need to be differenced to obtain stationarity and $q$ is the order of the moving average component. If we don't need to difference the series, we don't need to specify the "I" part and we can use the short version, i.e., the ARMA$(p,q)$. Consider three models: ARMA(1,0), ARMA(0,1) and ARMA(1,1) with parameters $\phi=0.6$ and $\theta= 0.9$. The $\phi$ refers to the AR coefficient and the $\theta$ refers to the MA coefficient. Use R to generate $n=100$ observations from each of these three models

```{r simulating ARMA models}

firstmodel <- arima.sim(model=list(order=c(1,0,0), ar=0.6), n=100)
firstmodel

secondmodel <- arima.sim(model=list(order=c(0,0,1), ma = 0.9), n=100)
secondmodel

thirdmodel <- arima.sim(model=list(order=c(1,0,1), ar=0.6, ma = 0.9), n=100)
thirdmodel
```


\begin{enumerate}[label=(\alph*)]

\item Plot the sample ACF for each of these models in one window to facilitate comparison (Hint: use command $par(mfrow=c(1,3))$ that divides the plotting window in three columns).  

```{r plotting ACFs}

par(mfrow=c(1,3))

acf(firstmodel, lag.max = 40)
acf(secondmodel, lag.max = 40)
acf(thirdmodel, lag.max = 40)


```


\item Plot the sample PACF for each of these models in one window to facilitate comparison.  

```{r plotting PACFs}

par(mfrow=c(1,3))

pacf(firstmodel, lag.max = 40)
pacf(secondmodel, lag.max = 40)
pacf(thirdmodel, lag.max = 40)



```

\item Look at the ACFs and PACFs. Imagine you had these plots for a data set and you were asked to identify the model, i.e., is it AR, MA or ARMA and the order of each component. Would you be identify them correctly? Explain your answer.

> Answer: It is possible to distinguish between the models by looking at their ACFs and PACFs in the following ways:
1) The PACF of Model 1 does not decay gradually but its ACF does gradually decay which shows that it is an AR series. Hence, its order can be checked by looking at its PACF cut off which happens at lag 1, and hence its order is 1.
2) The PACF of Model 2 decays gradually which shows a MA component in the series. Its order can be checked by its ACF plot which shows a cutoff after lag 2, and hence its order is 2.
3) The PACF of Model 3 shows a gradual decay in both ACF and PACF plots which shows the presence of both AR and MA components. It is unclear on how to determine its order by looking at the plots.

\item Compare the ACF and PACF values R computed with the theoretical values you provided for the coefficients. Do they match? Explain your answer.

> Answer: For Model-1, the PACF lag 1 cuts off at 0.6 which is equal to the AR coefficient that was entered in.


\item Increase number of observations to $n=1000$ and repeat parts (a)-(d).

```{r increasing n to 1000}
firstmodel2 <- arima.sim(model=list(order=c(1,0,0), ar=0.6), n=1000)
secondmodel2 <- arima.sim(model=list(order=c(0,0,1), ma = 0.9), n=1000)
thirdmodel2 <- arima.sim(model=list(order=c(1,0,1), ar=0.6, ma = 0.9), n=1000)


par(mfrow=c(1,3))

plot(firstmodel2)
plot(secondmodel2)
plot(thirdmodel2)

par(mfrow=c(1,3))

pacf(firstmodel2, lag.max = 40)
pacf(secondmodel2, lag.max = 40)
pacf(thirdmodel2, lag.max = 40)


par(mfrow=c(1,3))

acf(firstmodel2, lag.max = 40)
acf(secondmodel2, lag.max = 40)
acf(thirdmodel2, lag.max = 40)

```

\item Look at the ACFs and PACFs. Imagine you had these plots for a data set and you were asked to identify the model, i.e., is it AR, MA or ARMA and the order of each component. Would you be identify them correctly? Explain your answer.

> Answer: It is even more easy to distinguish between the AR and MA models because the pace of decay of the plots is much more evident now than it was with n=100. So the results from n=100 are felt even stronger in this iteration. 


\item Compare the ACF and PACF values R computed with the theoretical values you provided for the coefficients. Do they match? Explain your answer.

> Answer: For Model-1, the PACF lag 1 cuts off at near 0.6 which is equal to the AR coefficient that was entered in.

\end{enumerate}

## Q3

Consider the ARIMA model $y_t=0.7*y_{t-1}-0.25*y_{t-12}+a_t-0.1*a_{t-1}$

\begin{enumerate}[label=(\alph*)]

\item Identify the model using the notation ARIMA$(p,d,q)(P,D,Q)_ s$, i.e., identify the integers $p,d,q,P,D,Q,s$ (if possible) from the equation.

\item Also from the equation what are the values of the parameters, i.e., model coefficients. 

> Answer: SARIMA(1,0,1)(1,0,0)[12] SAR term = -0.25, phi=0.7, theta=0.1



\end{enumerate}
## Q4

Plot the ACF and PACF of a seasonal ARIMA$(0, 1)\times(1, 0)_{12}$ model with $\phi =0 .8$ and $\theta = 0.5$ using R. The $12$ after the bracket tells you that $s=12$, i.e., the seasonal lag is 12, suggesting monthly data whose behavior is repeated every 12 months. You can generate as many observations as you like. Note the Integrated part was omitted. It means the series do not need differencing, therefore $d=D=0$. Plot ACF and PACF for the simulated data. Comment if the plots are well representing the model you simulated, i.e., would you be able to identify the order of both non-seasonal and seasonal components from the plots? Explain.

```{r SARIMA simulation}

sarimasimul<- sim_sarima(n=1000, model=list(ar=0, ma=1, sar=1, sma=0), S=12)
sarimasimul

par(mfrow=c(1,2))
acf(sarimasimul, lag.max = 40)
pacf(sarimasimul, lag.max = 40)

```


> Answer: I can identify the model that I simulated because I can verify the non-seasonal and seasonal order and behavior using the ACF and PACF plots. Here, the ACF cuts off which shows that it is an MA process with order 1 because of the cut-off spike at lag 1. Similarly, the PACF shows gradual decay which reinforces this claim.
For the seasonal component, the multiple spikes at 12, 24 and 36 show that it is an SAR process, and the single spike in the PACF reinforces this claim. 
